---
title: "Final"
author: Jocelyn Magner
date: 6-4-26
format: pdf
---
# Set up
[Github repo:](https://github.com/jmagner4/ENVS-193DS_spring-2026_final)

## Read in packages
```{r}
#| label: Set Up
#| message: false
#| warning: false
# read in packages
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection
library(DHARMa) # diagnostic plots

# store data in new object
nests<- read_csv(here("data","nest_data_final.csv"))
```

# Problems
## Problem 1. Research Writing
### a.Transparent statistical methods

In part 1, I think they used binomial/logistic regression. If habitat use is measured as used/not used, then the response variable would be binary.
In part 2, I think they used a Spearman's correlation test. Wanting to know the relationship between two variables suggests a correlation test should be used. Since both variables are categorical/nominal that suggests a non-parametric test should be used.

### b. Suggetions for Rewriting

Part 1: Our evidence suggests that the distance to the water's edge is an important factor in determining where American Avocet choose to live.

Part 2: Different shorebird species, including American Avocet, Forster’s Tern, Black-necked Stilt, prefer different types of coastal habitats.

### c. Further Analysis

One potential variable influencing the probability of American Avocet microhabitat use is the availability of food or prey species in the habitats as well; this population would be a discrete variable (measured as a number of individuals per one-meter net sweep). I expect American Avocets to choose areas with higher prey populations because it makes finding food and surviving easier.

## Problem 2. Data Analysis

### a. Clean and Wrangle Data
```{r}
# create new object for cleaned data
nests_clean <- nests |>
  
# select columns of interest
  select(ant_species, case_control, height_cm)|>

# include only 3 ants species of interest
  filter(ant_species %in% c("AB", "BBR", "RRB"))|>

# factor reorders levels of ant species as directed 
  mutate(
    ant_species = factor(ant_species, levels = c("AB", "RRB", "BBR")),
    
# include new column of full scientific names       
    full_scientific_name = case_when(
      ant_species == "AB" ~ "Crematogaster sjostedti",
      ant_species == "RRB" ~ "Crematogaster mimosae",
      ant_species == "BBR" ~ "Crematogaster nigriceps"))

# display structure
str(nests_clean)

# display 10 random samples
nests_clean |> slice_sample(n = 10)
```

### b. Response Variable

For the case_control column, a 1 represents a tree where a bird nest is present while a 0 represents an "available" tree where no birds nests are present.

### c. Purpose of Study

Tree height is a continuous variable. As tree height increases the probability of nest occurrence would increase according to paragraph 2 of the results section. The authors say tree architecture has an important role in nest site selection in the second paragraph of the discussion; denser tree cover likely provides better nest concealment, protecting bird eggs from predators (but nothing about height specifically here).


Acacia ant species is a categorical variable (confirmed in the methods section). The probability of finding a bird's nest is higher for trees occupied by C. nigriceps compared to C. mimosae (found in second paragraph of results section). 


### d. Table of Models

4 models total:

| Model number | height | ant species | Model Description                      |
|-------------:|:------:|:-----------:|:--------------------------------------:|
| 1            |        |             | no interactions (no predictors present)|
| 2            |    X   |      X      | interaction (height and ant species)   | 

### e. Run the Models

```{r}
#| label: model fitting
#| include: false

# model 1: no interaction occurs
# use glm because we want to know how a variable "predicts" or "affects" another and the response variable is non-parametric
# add family = binomial because the response variable is binary 
model1 <-glm(case_control ~ ant_species + height_cm,
            data = nests_clean,
            family = "binomial")

# model 2: interaction
model2 <-glm(case_control ~ ant_species * height_cm,
            data = nests_clean,
            family = "binomial")

```

### f. Select the Best Model

```{r}
#| label: Model Selection
AICc(model1, model2)|>
  arrange(-AICc)
```
The best model as determined by Akaike's Information Criterion (AIC) is model 1. This model suggests there is an interactive effect of ant species and tree height on bird nest occurrence.

### g. Check the Diagnostics

```{r}
#| label: check diagnostics

plot(simulateResiduals(model2))
```

### h. Visualize the Model

```{r}
# generate predictions to plot
mod_preds <- ggpredict(model2, terms = c("height_cm [all]", "ant_species"))
```

```{r}

# base layer ggplot
ggplot() +

  # Raw data points 
  geom_point(data = nests_clean, 
             aes(x = height_cm, 
                 y = case_control, 
                 color = ant_species, 
                 group = ant_species,
             alpha = 0.3, size = 2.5)) +
  
  # 95% confidence interval ribbons
  geom_ribbon(data = mod_preds, 
              aes(x = x, y = predicted, ymin = conf.low, ymax = conf.high, fill = group), 
              alpha = 0.15, inherit.aes = FALSE) +
  
  # Logistic regression prediction lines
  geom_line(data = mod_preds, 
            aes(x = x, y = predicted, color = group), 
            linewidth = 1.2, inherit.aes = FALSE) +
  
  # Scale and presentation formatting
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  
  # display model predictions for each ant species in a different panel
  facet_wrap(~group) + 
  
  # Take out grid lines and gray background
  theme_minimal() +
    theme(panel.grid = element_blank(), # Removes grid
    text = element_text(face = "bold", size = 11))+

  # Use colors that are different from the default
  scale_color_manual(values = c("#FEF65B","#FF2088","#0041C2")) +
  scale_fill_manual(values = c("#FEF65B","#FF2088", "#0041C2")) +

  # Add labels with units
  labs(
    x = "Height (cm)",
    y = "Probability of Bird Nest Presence",
    color = "Ant Species",
    fill = "Ant Species")
```


### i. Write a caption for your figure

**Figure 1. The probability of bird nest presence differs depending on tree height and ant species present.** The circles represent raw binary observations of bird nest presence (1.0) or absence (0.0) at a specific tree height. The colors represent different ant species occupation of the trees sampled; yellow represents Crematogaster sjostedti (AB), pink represents Crematogaster mimosae (RRB), blue represents Crematogaster nigriceps (BBR). The lines represent the predicted probability of a bird nesting at that corresponding height. The translucent ribbons represent 95% confidence intervals for each species represented. Data from: Lujan, E., Nielsen, R., Short, Z., Wicks, S., Nderitu Watetu, W., Malingati Khasoha, L., Palmer, T. M., Goheen, J. R., & Alston, J. M. (2023). Data, code, and metadata for: Symbiotic acacia ants drive nesting behavior by birds in an African savanna [Data set]. Zenodo. https://doi.org/10.5281/zenodo.8373322

### j. Calculate Model Predictions

```{r}
#| echo: true
# calculate model predictions with model 2 since it was the least bad accordign to AICc
ggpredict(model2,
          terms = c("height_cm [600]", # include height at 600cm as instructed
                    "ant_species")) # include ant species also
```


### k. Interpret Your Results

Ant species and tree height interact to predict the probability of bird nest occurrence. 

Birds are more likely to nest in trees with aggressive defending ant species like C. nigriceps (BBR) and C. mimosae (RRB), likely because these ants ward off predators. Birds also prefer to nest in taller trees likely because it is harder for land predators to reach their eggs.

Additionally the presence of C. nigriceps effects tree structure making birds most likely to choose these trees. In figure 1 the slope of the blue line is steep at ~150 cm suggesting a high probability that birds will nest in BBR trees taller than 150 cm.

In the tallest tree (600 cm), the probability of nesting is 0.20 (95% CI:[0.00,0.99]) for AB, 0.74 (95% CI:[0.48,0.90]) for RRB, and 1.00 (95% CI:[1.00,1.00]) for BBR.



## Problem 3. Affective and explanatory visualizations

### a.
### b.
### c.
### d.
### e.
### f.
### g.