---
title: "Final"
author: Jocelyn Magner
date: 6-4-26
format: pdf
---
# Set up
[Github repo:](https://github.com/jmagner4/ENVS-193DS_spring-2026_final)

## Read in packages
```{r}
#| label: Set Up
#| message: false
#| warning: false
# read in packages
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# store data in new object
nests<- read_csv(here("data","nest_data_final.csv"))
```

# Problems
## Problem 1. Research Writing
### a.Transparent statistical methods

In part 1, I think they used binomial/logistic regression. If habitat use is measured as used/not used, then the response variable would be binary.
In part 2, I think they used a Spearman's correlation test. Wanting to know the relationship between two variables suggests a correlation test should be used. Since both variables are categorical/nominal that suggests a non-parametric test should be used.

### b. Suggetions for Rewriting

Part 1: Our evidence suggests that the distance to the water's edge is an important factor in determining where American Avocet choose to live.

Part 2: Different shorebird species, including American Avocet, Forster’s Tern, Black-necked Stilt, prefer different types of coastal habitats.

### c. Further Analysis

One potential variable influencing the probability of American Avocet microhabitat use is the availability of food or prey species in the habitats as well; this population would be a discrete variable (measured as a number of individuals per one-meter net sweep). I expect American Avocets to choose areas with higher prey populations because it makes finding food and surviving easier.

## Problem 2. Data Analysis

### a. Clean and Wrangle Data
```{r}
# create new object for cleaned data
nests_clean <- nests |>
  
# select columns of interest
  select(ant_species, case_control, height_cm)|>

# include only 3 ants species of interest
  filter(ant_species %in% c("AB", "BBR", "RRB"))|>

# factor reorders levels of ant species as directed 
  mutate(
    ant_species = factor(ant_species, levels = c("AB", "RRB", "BBR")),
    
# include new column of full scientific names       
    full_scientific_name = case_when(
      ant_species == "AB" ~ "Crematogaster sjostedti",
      ant_species == "RRB" ~ "Crematogaster mimosae",
      ant_species == "BBR" ~ "Crematogaster nigriceps"))

# display structure
str(nests_clean)

# display 10 random samples
nests_clean |> slice_sample(n = 10)
```

### b. Response Variable

For the case_control column, a 1 represents a tree where a bird nest is present while a 0 represents an "available" tree where no birds nests are present.

### c. Purpose of Study

Tree height is a continuous variable. As tree height increases the probability of nest occurrence would increase according to paragraph 2 of the results section. It is unclear specifically why, but they do say tree architecture has an important role in nest site selection in the discussion; maybe because tall trees have less predators.


Acacia ant species is a categorical variable (confirmed in the methods section). The probability of finding a bird's nest is higher for trees occupied by C. nigriceps compared to C. mimosae (found in second paragraph of results section). 


### d. Table of Models

4 models total:

| Model number | height | ant species | Predictor list              |
|-------------:|:------:|:-----------:|:---------------------------:|
| 1            |        |             | no interactions             |
| 2            |    X   |      X      | interaction                 | 

### e. Run the Models

```{r}
#| label: model fitting
#| include: false

# model 1: no interaction occurs
model1 <-lm(case_control ~ height_cm + ant_species,
            data = nests_clean)

# model 2: interaction
model2 <-lm(case_control ~ height_cm * ant_species,
            data = nests_clean)


```

### f. Select the Best Model

```{r}
#| label: Model Selection
AICc(model1, model2)|>
  arrange(-AICc)
```


### g. Check the Diagnostics

```{r}
#| label: check diagnostics

par(mfrow = c(2,2))
    plot(model1)

```

### h. Visualize the Model
  

### i. Write a caption for your figure

### j. Calculate Model Predictions

### k. Interpret Your Results

## Problem 3. Affective and explanatory visualizations
### a.
### b.
### c.
### d.
### e.
### f.
### g.